A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Finished editing sections Data and Methods and improved the plots for the results

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive. For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test. The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project [**by whom?**]. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., AEMET, the state meteorology agency, and MAPA, the ministry of agriculture, fisheries, and food). The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute). Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.

@@

```{r descriptive-statistics, echo=FALSE, cache=TRUE}
data.frame(Variable = c("COVID-19 Incidence",
                        "Area", 
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "Area of province in sq.km",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                    mean(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                    mean(provinces_spain$GDPpc),
                    mean(provinces_spain$Older),
                    mean(provinces_spain$Density),
                    mean(covid19_spain$Mean_Temp),
                    mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                  sd(provinces_spain$GDPpc),
                  sd(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                  sd(provinces_spain$Older),
                  sd(provinces_spain$Density),
                  sd(covid19_spain$Mean_Temp),
                  sd(covid19_spain$Humidity)),
           Source = c("Montera34",
                      "INE",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable("latex",
        #"html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics} Descriptive statistics") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  #kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("Montera34: https://code.montera34.com:4443/numeroteca/covid19/-/blob/master/README_providencialdata19.md", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

@@

where $\bf{Y}$ is now a vector of dimension $NT \times 1$, $\bf{X}$ is a block-diagonal matrix $NT \times K$ 
(with $K = \sum_t{k_t}$) and $\epsilon$ is a $NT \times 1$ vector. Using the Kronecker product notation ($\otimes$), the error matrix structure is expressed concisely as:

@@

<!--
  ![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)
-->

@@

The stack expresion for the spatial lag SUR model (SUR-SLM) is as follows:

@@

\noindent where $\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$ with $\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$.

<!--
  ![\begin{equation} \label{eq:sur-slm} \bf{A}y = X \beta + \epsilon \\ \epsilon =N(0,\Omega) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbf%7BA%7Dy%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5COmega)%20%5Cend%7Bequation%7D)

\noindent where ![$A = I_{TN}-\bf{\Lambda} \otimes W$ with $\bf{\Lambda} = diag(\lambda_1, \cdots, \lambda_T)$](https://render.githubusercontent.com/render/math?math=%24A%20%3D%20I_%7BTN%7D-%5Cbf%7B%5CLambda%7D%20%5Cotimes%20W%24%20with%20%24%5Cbf%7B%5CLambda%7D%20%3D%20diag(%5Clambda_1%2C%20%5Ccdots%2C%20%5Clambda_T)%24).
-->

This specification assumes that incidence in a province ($y_{st}$) at time $t$ is partially determined by the weighted average ($Wy_{st}$) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period $\rho_t$ and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypothesis of identtical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$). The correspond Wald test is available in the `R` package `spsur`.

<!--
This specification assumes that incidence in a province (![$y_st$](https://render.githubusercontent.com/render/math?math=%24y_st%24)) at time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average ![($Wy_{st}$)](https://render.githubusercontent.com/render/math?math=(%24Wy_%7Bst%7D%24)) of incidence in neighbouring provinces. Coefficients for the spatially lagged variable are estimated for each time period ![$\lambda_t$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_t%24) and identifies the intensity and the sign of the impacts of neighbourhood. It is possible test the null hypotheis of identtical levels of spatial dependence (![$\lambda_i=\lambda_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Clambda_i%3D%5Clambda_j%2C%20%5Cforall%20i%2Cj%24)). The correspond Wald test is available in the `R` package `spsur`. 
-->

@@

```{r daily-correlations, echo=FALSE, fig.cap="\\label{fig:daily-correlations}Distribution of daily correlations of the independent variables with daily incidence of COVID-19 (all variables have been log-transformed)", cache = TRUE}
covid19_spain %>% 
  group_by(Date) %>%
  summarize(Older = cor(log(Older), log(Incidence)),
            GDPpc = cor(log(GDPpc), log(Incidence)),
            Density = cor(log(Density), log(Incidence)),
            Humidity_lag8 = cor(Humidity_lag8, Incidence),
            Humidity_lag11 = cor(log(Humidity_lag11), log(Incidence)),
            Humidity_lag11w = cor(log(Humidity_lag11w), log(Incidence)),
            Mean_Temp_lag8 = cor(log(Mean_Temp_lag8), log(Incidence)),
            Mean_Temp_lag11 = cor(log(Mean_Temp_lag11), log(Incidence)),
            Mean_Temp_lag11w = cor(log(Mean_Temp_lag11w), log(Incidence))) %>%
  select(-Date) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  mutate(Variable = factor(Variable,
                           levels = c("GDPpc", 
                                      "Older", 
                                      "Density", 
                                      "Humidity_lag8",
                                      "Humidity_lag11",
                                      "Humidity_lag11w",
                                      "Mean_Temp_lag8",
                                      "Mean_Temp_lag11",
                                      "Mean_Temp_lag11w"))) %>%
  ggplot(aes(x = Variable, y = Correlation)) +
  geom_boxplot() +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90))
```

@@

```{r delta-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of spatial autocorrelation coefficient"}
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Deltas = sur.slm_lag11$deltas,
           tvalue = sur.slm_lag11$deltas/sur.slm_lag11$deltas.se) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Deltas, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

@@

```{r beta-intercept-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept"}
#Intercept
n = 1
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

@@

```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density)", cache = TRUE}
#Density:
n = 2 # Although this is the fourth coefficient, after intercept, log(GDPpc), and log(Older), we start at 2 here because the two constrained coefficients are removed from the matrix of coefficients 
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

@@

```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit", cache = TRUE}
n = 3
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

@@

```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity)", cache = TRUE}
n = 4
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

@@

```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature)", cache = TRUE}
n = 5
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = matrix(sur.slm_lag11$coefficients[-c(2,3)],ncol = T)[n,],
           tvalue = matrix(sur.slm_lag11$coefficients[-c(2,3)]/sur.slm_lag11$rest.se[-c(2,3)], ncol = T)[n,]) %>%
  mutate(sig = case_when(abs(tvalue) >= qnorm(0.95, mean = 0, sd = 1) ~ "p <= 0.05",
                         abs(tvalue) >= qnorm(0.90, mean = 0, sd = 1) ~ "p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = sig)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "black")) +  geom_hline(yintercept = 0) +
  geom_vline(xintercept = as_date("2020-03-16"), color = "red") +
  theme_tufte()
```

@@

myimpact <- matrix(0,ncol = 5, nrow = 29)
for (i in 2:30){
  myimpact[i-1,] <- impacts.sur.slm_lag11[[i]]$res$direct
}
matplot(myimpact, type='l')

>>>>

Our dataset includes information about the daily number of cases of COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat terminology) for the period between March 13th and April 11th, inclusive. For our purposes, we consider positive cases reported, but excluding symptomatic cases diagnosed by a doctor without a Polymerase Chain Reaction (PCR) test. The Spanish National Government publishes periodic updates at the regional level (NUTII) and the information is also released at the provincial level as part of a collaborative project by geovoluntarios.com\footnote{https://www.geovoluntarios.org/}, ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19}, and ESRI España. This information is compiled from several sources, mainly the official web pages of the Spanish regional goverments, as documented in Centro de Datos Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}. In addition, we consider two sets of explanatory variables. The first one, and the focus of this research, is set of two environmental variables, namely temperature and humidity, which are collected from official sources (i.g., *AEMET*, the state meteorology agency, and *MAPA*, the ministry of agriculture, fisheries, and food). The second set provides some relevant controls for multivariate analysis, and refers to economic and demographic attributes of the province (also collected from official sources, i.e., INE, the national statistics institute). Table \ref{tab:descriptive-statistics} shows the descriptive statistics and the provenance of the data.

@@

```{r descriptive-statistics, echo=FALSE, cache=TRUE}
data.frame(Variable = c("COVID-19 Incidence",
                        "Area", 
                        "GDPpc", 
                        "Older", 
                        "Population Density",
                        "Mean Temperature",
                        "Humidity"),
           Note = c("Incidence in reported cases of SARS-19 per 100,000 people",
                    "Area of province in sq.km",
                    "GDP per capita in €1,000s",
                    "Percentage of people aged 65 and older in the province",
                    "Population density in the province in people per sq.km",
                    "Mean temperature in province by date, in C",
                    "Relative humidity in province by date"),
           Min = c(min(covid19_spain$Incidence),
                   min(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   min(provinces_spain$GDPpc),
                   min(provinces_spain$Older),
                   min(provinces_spain$Density),
                   min(covid19_spain$Mean_Temp),
                   min(covid19_spain$Humidity)),
           Mean = c(mean(covid19_spain$Incidence),
                    mean(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                    mean(provinces_spain$GDPpc),
                    mean(provinces_spain$Older),
                    mean(provinces_spain$Density),
                    mean(covid19_spain$Mean_Temp),
                    mean(covid19_spain$Humidity)),
           Max = c(max(covid19_spain$Incidence),
                   max(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                   max(provinces_spain$GDPpc),
                   max(provinces_spain$Older),
                   max(provinces_spain$Density),
                   max(covid19_spain$Mean_Temp),
                   max(covid19_spain$Humidity)),
           SD = c(sd(covid19_spain$Incidence),
                  sd(provinces_spain$GDPpc),
                  sd(provinces_spain %>% 
                         st_area() %>%
                         set_units(km^2)),
                  sd(provinces_spain$Older),
                  sd(provinces_spain$Density),
                  sd(covid19_spain$Mean_Temp),
                  sd(covid19_spain$Humidity)),
           Source = c("ProvidencialData19",
                      "INE",
                      "INE",
                      "INE",
                      "INE",
                      "AEMET",
                      "MAPA")) %>%
  kable("latex",
        #"html",
        booktabs = TRUE,
        digits = 2,
        caption = "\\label{tab:descriptive-statistics} Descriptive statistics") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  #kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(7, width = "5em") %>%
  footnote(general = c("ProvidencialData19: https://www.datoscovid.es/pages/providencialdata19", 
                       "INE (Instituto Nacional de Estadistica): https://www.ine.es/",
                       "AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es",
                       "MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es"))
```

@@

\noindent where $\bf{Y}$ is now a vector of dimension $NT \times 1$, $\bf{X}$ is a block-diagonal matrix $NT \times K$ 
(with $K = \sum_t{k_t}$) and $\epsilon$ is an $NT \times 1$ vector. Using the Kronecker product notation ($\otimes$), the error matrix structure is expressed concisely as:

@@

<!--
![\begin{equation} y = X \beta + \epsilon \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20y%20%3D%20X%20%5Cbeta%20%2B%20%5Cepsilon%20%5Cend%7Bequation%7D)

\noindent where ![$\bf{Y}$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BY%7D%24) is now a vector of dimension ![$NT \times 1$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%201%24), ![$\bf{X}$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BX%7D%24) is a block-diagonal matrix ![$NT \times K$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%20K%24) 
(with ![$K = \sum_t{k_t}$](https://render.githubusercontent.com/render/math?math=%24K%20%3D%20%5Csum_t%7Bk_t%7D%24)) and ![$\epsilon$](https://render.githubusercontent.com/render/math?math=%24%5Cepsilon%24) is an ![$NT \times 1$](https://render.githubusercontent.com/render/math?math=%24NT%20%5Ctimes%201%24) vector. Using the Kronecker product notation (![$\otimes$](https://render.githubusercontent.com/render/math?math=%24%5Cotimes%24)), the error matrix structure is expressed concisely as:

![\begin{equation} E\[\epsilon \epsilon'\]=\Sigma \otimes I_N ; \ \Sigma=(\sigma_{tt'}) \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20E%5B%5Cepsilon%20%5Cepsilon'%5D%3D%5CSigma%20%5Cotimes%20I_N%20%3B%20%5C%20%5CSigma%3D(%5Csigma_%7Btt'%7D)%20%5Cend%7Bequation%7D)
-->

@@

The stack expresion for the SUR model with a spatially lagged dependent variable (SUR-SLM) is as follows:

@@

\noindent where $\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$ is the spatially lagged dependent variable, and $\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$.

<!--
![\begin{equation} \label{eq:sur-slm} \begin{aligned} \bf{AY} = \bf{X} \beta + \epsilon \\ \epsilon =N(0,\bf{\Sigma}) \end{aligned} \end{equation}](https://render.githubusercontent.com/render/math?math=%5Cbegin%7Bequation%7D%20%5Clabel%7Beq%3Asur-slm%7D%20%5Cbegin%7Baligned%7D%20%5Cbf%7BAY%7D%20%3D%20%5Cbf%7BX%7D%20%5Cbeta%20%2B%20%5Cepsilon%20%5C%5C%20%5Cepsilon%20%3DN(0%2C%5Cbf%7B%5CSigma%7D)%20%5Cend%7Baligned%7D%20%5Cend%7Bequation%7D)

\noindent where ![$\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BA%7D%20%3DI_%7BTN%7D-%5Cbf%7B%5CGamma%7D%20%5Cotimes%20W%24) is the spatially lagged dependent variable, and ![$\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7B%5CGamma%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
-->

This specification assumes that outcome ($y_{st}$) at location $s$ and time $t$ is partially determined by the weighted average ($Wy_{st}$) of the outcome in neighboring provinces, with neighborhood defined by matrix $W$ of spatial weights. In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way. The coefficients of the spatially lagged variable are estimated for each time period $\rho_t$ and identify the intensity and the sign of the contagion effect. It is possible test the null hypothesis of identical levels of spatial dependence ($\rho_i=\rho_j, \forall i,j$). The correspond Wald test is available in the `R` package `spsur`.

<!--
This specification assumes that outcome (![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24)) at location ![$s$](https://render.githubusercontent.com/render/math?math=%24s%24) and time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average (![$Wy_{st}$](https://render.githubusercontent.com/render/math?math=%24Wy_%7Bst%7D%24)) of the outcome in neighboring provinces, with neighborhood defined by matrix ![$W$](https://render.githubusercontent.com/render/math?math=%24W%24) of spatial weights. In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way. The coefficients of the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identify the intensity and the sign of the contagion effect. It is possible test the null hypothesis of identical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)). The correspond Wald test is available in the `R` package `spsur`. 
-->

@@

```{r daily-correlations, echo=FALSE, fig.cap="\\label{fig:daily-correlations}Distribution of daily correlations of the independent variables with daily incidence of COVID-19 (all variables have been log-transformed)", cache = TRUE}
covid19_spain %>% 
  group_by(Date) %>%
  summarize(Older = cor(log(Older), log(Incidence)),
            GDPpc = cor(log(GDPpc), log(Incidence)),
            Density = cor(log(Density), log(Incidence)),
            Humidity_lag8 = cor(Humidity_lag8, Incidence),
            Humidity_lag11 = cor(log(Humidity_lag11), log(Incidence)),
            Humidity_lag11w = cor(log(Humidity_lag11w), log(Incidence)),
            Mean_Temp_lag8 = cor(log(Mean_Temp_lag8), log(Incidence)),
            Mean_Temp_lag11 = cor(log(Mean_Temp_lag11), log(Incidence)),
            Mean_Temp_lag11w = cor(log(Mean_Temp_lag11w), log(Incidence)),
            Sunshine_Hours_lag8 = cor(log(Sunshine_Hours_lag8 + 0.1), log(Incidence)),
            Sunshine_Hours_lag11 = cor(log(Sunshine_Hours_lag11 + 0.1), log(Incidence)),
            Sunshine_Hours_lag11w = cor(log(Sunshine_Hours_lag11w + 0.1), log(Incidence))) %>%
  select(-Date) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Correlation") %>%
  mutate(Variable = factor(Variable,
                           levels = c("GDPpc", 
                                      "Older", 
                                      "Density", 
                                      "Humidity_lag8",
                                      "Humidity_lag11",
                                      "Humidity_lag11w",
                                      "Mean_Temp_lag8",
                                      "Mean_Temp_lag11",
                                      "Mean_Temp_lag11w",
                                      "Sunshine_Hours_lag8",
                                      "Sunshine_Hours_lag11",
                                      "Sunshine_Hours_lag11w"))) %>%
  ggplot(aes(x = Variable, y = Correlation)) +
  geom_boxplot() +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90))
```

@@

```{r delta-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:delta-time}Temporal evolution of the spatial autocorrelation coefficient; dots are the point estimates and vertical lines are 95% confidence intervals. In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed."}
data.frame(Date = seq(ymd("2020-03-13"), 
                      ymd("2020-04-11"), 
                      by = "days"),
           Beta = sur.slm_lag11$deltas,
           sd = sur.slm_lag11$deltas.se) %>%
  mutate(zscore = Beta/sd,
         pval = 2 * pnorm(-abs(zscore)),
           CI95low = Beta - qnorm(0.975, mean = 0, sd = 1) * sd,
           CI95up = Beta + qnorm(0.975, mean = 0, sd = 1) * sd,
           CI90low = Beta - qnorm(0.95, mean = 0, sd = 1) * sd,
           CI90up = Beta + qnorm(0.95, mean = 0, sd = 1) * sd,
         pval = case_when(pval <= 0.05 ~ "p <= 0.05",
                          pval > 0.05 & pval <= 0.10 ~ "0.05 < p <= 0.10",
                         TRUE ~ "p > 0.10")) %>%
  ggplot(aes(x = Date, y = Beta, color = pval)) +
  geom_rect(xmin = as_date("2020-03-14"), xmax = as_date("2020-03-30"),
              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
  geom_rect(xmin = as_date("2020-03-30"), xmax = as_date("2020-04-12"),
            ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
  geom_point(size = 3) +
  geom_errorbar(aes(x = Date,
                     ymin = CI90low,
                    ymax = CI90up), 
                width = 0.5) +
  geom_linerange(aes(x = Date,
                     ymin = CI95low,
                    ymax = CI95up),
                 size = 0.5) +
  scale_color_manual(values = c("0.05 < p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "gray")) +  
  geom_hline(yintercept = 0) +
  theme_bw()
```
```{r function-for-plotting-coefficients, include=FALSE}
# Define a function for plotting the coefficients over time including their confidence intervals and significance
plot_coeff <- function(model = model, n = n){
  data.frame(Date = seq(ymd("2020-03-13"), 
                        ymd("2020-04-11"), 
                        by = "days"),
             Beta = matrix(model$coefficients[-c(2,3)],ncol = T)[n,],
             sd = matrix(model$rest.se[-c(2,3)], ncol = T)[n,]) %>%
    mutate(zscore = Beta/sd,
           pval = 2 * pnorm(-abs(zscore)),
           CI95low = Beta - qnorm(0.975, mean = 0, sd = 1) * sd,
           CI95up = Beta + qnorm(0.975, mean = 0, sd = 1) * sd,
           CI90low = Beta - qnorm(0.95, mean = 0, sd = 1) * sd,
           CI90up = Beta + qnorm(0.95, mean = 0, sd = 1) * sd,
           pval = case_when(pval <= 0.05 ~ "p <= 0.05",
                            pval > 0.05 & pval <= 0.10 ~ "0.05 < p <= 0.10",
                            TRUE ~ "p > 0.10")) %>%
    ggplot(aes(x = Date, y = Beta, color = pval)) +
    geom_rect(xmin = as_date("2020-03-14"), xmax = as_date("2020-03-30"),
              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "yellow", color = NA) +
    geom_rect(xmin = as_date("2020-03-30"), xmax = as_date("2020-04-12"),
              ymin = -Inf, ymax = Inf, alpha = 0.010, fill = "orange", color = NA) +
    geom_point(size = 3) +
    geom_errorbar(aes(x = Date,
                      ymin = CI90low,
                      ymax = CI90up), 
                  width = 0.5) +
    geom_linerange(aes(x = Date,
                       ymin = CI95low,
                       ymax = CI95up),
                   size = 0.5) +
    scale_color_manual(values = c("0.05 < p <= 0.10" = "orange", "p <= 0.05" = "red", "p > 0.10" = "gray")) +  
    geom_hline(yintercept = 0) +
    theme_bw()
  }
```

@@

```{r beta-intercept-time, cache = TRUE, echo=FALSE, fig.cap="\\label{fig:beta-intercept-time}Temporal evolution of intercept; dots are the point estimates and vertical lines are 95% confidence intervals. In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed."}
#Intercept
plot_coeff(model = sur.slm_lag11, n = 1) + 
  ylab("Beta (Intercept)")
```

@@

```{r beta-density-time, echo=FALSE, , fig.cap="\\label{fig:beta-density-time}Temporal evolution of coefficient for log(Density); dots are the point estimates and vertical lines are 95% confidence intervals. In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
#Density:
plot_coeff(model = sur.slm_lag11, n = 2) + 
  ylab("Beta (log(Density))")
```

@@

```{r beta-transit-time, echo=FALSE, fig.cap="\\label{fig:beta-transit-time}Temporal evolution of coefficient for Transit; dots are the point estimates and vertical lines are 95% confidence intervals. In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
# Transit
plot_coeff(model = sur.slm_lag11, n = 3) + 
  ylab("Beta (Transit)")
```

@@

```{r beta-humidity-time, echo=FALSE, fig.cap="\\label{fig:beta-humidity-time}Temporal evolution of coefficient for log(Humidity); dots are the point estimates and vertical lines are 95% confidence intervals. In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
# Humidity:
plot_coeff(model = sur.slm_lag11, n = 4) + 
  ylab("Beta (log(Humidity))")
```

@@

```{r beta-temperature-time, echo=FALSE, fig.cap="\\label{fig:beta-temperature-time}Temporal evolution of coefficient for log(Temperature); dots are the point estimates and vertical lines are 95% confidence intervals. In yellow is the period after the declaration of the state of emergency, and in orange is the period when only essential activities were allowed.", cache = TRUE}
# Temperature
plot_coeff(model = sur.slm_lag11, n = 5) + 
  ylab("Beta (log(Temperature))")
```

@@

myimpact <- matrix(0,ncol = 5, nrow = 29)
for (i in 2:30){
  myimpact[i-1,] <- impacts.sur.slm_lag11[[i]]$res$direct
}

matplot(myimpact, type='l')
